---
title: "DATA 606 Data Project Proposal"
author: "Amber Ferger"
output: pdf_document
---

### Data Preparation

**First, I'll load in the data and rename the columns.**
```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}

library(dplyr)
library(tidyverse)

religionData <- read.csv('religionData.csv', header= TRUE, sep = ',')
religionData %>% as_tibble(religionData)

# rename columns
colNames <- c('RELIGION','RELIGION2', 'EVANGELICAL', 'RELIGIOUS_SERVICES', 'FREQ_PRAY_WITH_MOTIONS', 'FREQ_PRAY_WITH_OBJECTS', 'FREQ_PRAY_BEFORE_MEALS',
'FREQ_PRAY_FOR_OTHERS', 'FREQ_ASK_TO_PRAY_WITH_SOMEONE', 'FREQ_BRING_UP_RELIGION',
'FREQ_ASK_ABOUT_RELIGION', 'FREQ_DECLINE_FOOD_FOR_RELIGION', 'FREQ_WEAR_RELIGIOUS_CLOTHING', 'FREQ_PARTICIPATE_IN_PUBLIC_RELIGIOUS_EVENT',

'COMFORT_OWN_PRAY_WITH_MOTIONS',
'COMFORT_OWN_PRAY_WITH_OBJECTS',
'COMFORT_OWN_PRAY_BEFORE_MEALS',
'COMFORT_OWN_PRAY_FOR_OTHERS',
'COMFORT_OWN_ASK_TO_PRAY_WITH_SOMEONE',
'COMFORT_OWN_BRING_UP_RELIGION',
'COMFORT_OWN_ASK_ABOUT_RELIGION',
'COMFORT_OWN_DECLINE_FOOD_FOR_RELIGION',
'COMFORT_OWN_WEAR_RELIGIOUS_CLOTHING',
'COMFORT_OWN_PARTICIPATE_IN_PUBLIC_RELIGIOUS_EVENT',

'COMFORT_OTHER_PRAY_WITH_MOTIONS', 'COMFORT_OTHER_PRAY_WITH_OBJECTS', 'COMFORT_OTHER_PRAY_BEFORE_MEALS', 'COMFORT_OTHER_PRAY_FOR_OTHERS', 'COMFORT_OTHER_ASK_TO_PRAY_WITH_SOMEONE', 'COMFORT_OTHER_BRING_UP_RELIGION', 'COMFORT_OTHER_ASK_ABOUT_RELIGION', 'COMFORT_OTHER_DECLINE_FOOD_FOR_RELIGION',
'COMFORT_OTHER_WEAR_RELIGIOUS_CLOTHING', 'COMFORT_OTHER_PARTICIPATE_IN_PUBLIC_RELIGIOUS_EVENT', 'COMFORT_SEE_OTHER_PRAY_WITH_MOTIONS', 'COMFORT_SEE_OTHER_PRAY_WITH_OBJECTS', 'COMFORT_SEE_OTHER_PRAY_BEFORE_MEALS', 'COMFORT_SEE_OTHER_PRAY_FOR_OTHERS', 'COMFORT_SEE_OTHER_ASK_TO_PRAY_WITH_SOMEONE', 'COMFORT_SEE_OTHER_BRING_UP_RELIGION',
'COMFORT_SEE_OTHER_ASK_ABOUT_RELIGION', 'COMFORT_SEE_OTHER_DECLINE_FOOD_FOR_RELIGION', 'COMFORT_SEE_OTHER_WEAR_RELIGIOUS_CLOTHING', 'COMFORT_SEE_OTHER_PARTICIPATE_IN_PUBLIC_RELIGIOUS_EVENT', 'AGE', 'GENDER', 'HOUSEHOLD_SALARY', 'US_REGION')

names(religionData) <- colNames
```

**There are a number of fields within this dataset, so I am going to subset it only to general demographics and survey responses related to the comfort of seeing religious actions outside of the respondent's religion.**

```{r}
colsToKeep <- c('COMFORT_SEE_OTHER_PRAY_WITH_MOTIONS', 'COMFORT_SEE_OTHER_PRAY_WITH_OBJECTS', 'COMFORT_SEE_OTHER_PRAY_BEFORE_MEALS', 'COMFORT_SEE_OTHER_PRAY_FOR_OTHERS', 'COMFORT_SEE_OTHER_ASK_TO_PRAY_WITH_SOMEONE', 'COMFORT_SEE_OTHER_BRING_UP_RELIGION',
'COMFORT_SEE_OTHER_ASK_ABOUT_RELIGION', 'COMFORT_SEE_OTHER_DECLINE_FOOD_FOR_RELIGION', 'COMFORT_SEE_OTHER_WEAR_RELIGIOUS_CLOTHING', 'COMFORT_SEE_OTHER_PARTICIPATE_IN_PUBLIC_RELIGIOUS_EVENT', 'AGE', 'GENDER', 'HOUSEHOLD_SALARY', 'US_REGION')

religionData <- religionData %>%
  filter(RELIGION != 'Response') %>%
  select(colsToKeep)

```

**In order to quantify each individual's comfort with public religious displays, I need to convert my categorical features to numeric. First, I will rank each survey response in order of comfort (with Not at all comfortable having the lowest ranking and Extremely comfortable having the highest ranking)**
``` {r} 
comfortColumns <- seq(1:10)

for (i in comfortColumns) {
  religionData[[i]] <- factor(religionData[[i]], levels = c("", "Response", "Not at all comfortable", "Not so comfortable", "Somewhat comfortably", "Very comfortable", "Extremely comfortable"), ordered = TRUE)
  
}

# eliminate records where any of the survey responses are blank
toBeRemoved<-which(religionData$COMFORT_SEE_OTHER_PRAY_WITH_MOTIONS==""|religionData$COMFORT_SEE_OTHER_PRAY_WITH_OBJECTS==""|religionData$COMFORT_SEE_OTHER_PRAY_BEFORE_MEALS==""|religionData$COMFORT_SEE_OTHER_PRAY_FOR_OTHERS==""|religionData$COMFORT_SEE_OTHER_ASK_TO_PRAY_WITH_SOMEONE==""|religionData$COMFORT_SEE_OTHER_BRING_UP_RELIGION==""|religionData$COMFORT_SEE_OTHER_ASK_ABOUT_RELIGION==""|religionData$COMFORT_SEE_OTHER_DECLINE_FOOD_FOR_RELIGION==""|religionData$COMFORT_SEE_OTHER_WEAR_RELIGIOUS_CLOTHING==""|religionData$COMFORT_SEE_OTHER_PARTICIPATE_IN_PUBLIC_RELIGIOUS_EVENT=="")

religionData<-religionData[-toBeRemoved,]
```

**The survey responses are still categorical, so I need to convert them to numeric values. Since the columns are already ordered, the numerical values should maintain the hierarchy. I'll also need to compute an average ranking across survey responses. I'll create a new variable called AVERAGE_RANKING with this information.**
``` {r}
religionData$AVERAGE_RATING <- 0


# loop through all survey questions and convert each response to a number 
# add number to the AVERAGE_RATING column
for (i in comfortColumns) {
  religionData[[i]]<-as.numeric(religionData[[i]])
  religionData$AVERAGE_RATING <- religionData$AVERAGE_RATING + religionData[[i]]
}

# final average rating
religionData$AVERAGE_RATING <- religionData$AVERAGE_RATING/10

```


### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

Are age, income, and region predictive of an individual's comfort level with public religious displays?



### Cases 

**What are the cases, and how many are there?**

Each case represents a respondent's survey answers. There are a total of 979 cases in the dataset.



### Data collection 

**Describe the method of data collection.**

This data was collected using a SurveyMonkey poll, conducted between July 29 and August 1, 2016. The survey asked 661 respondents questions about public displays of religion.



### Type of study 

**What type of study is this (observational/experiment)?**

This is a survey, which is a type of observational study.



### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

This data came from: https://github.com/fivethirtyeight/data/tree/master/religion-survey.




### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**

The response variable is comfort level and it is numeric.



### Independent Variable

**You should have two independent variables, one quantitative and one qualitative.**

* Quantitative Variable - Age, Income 
* Qualitative Variable - Region



### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}

summary(religionData$AVERAGE_RATING)
table(religionData$AGE)
table(religionData$HOUSEHOLD_SALARY)
table(religionData$US_REGION)

religionData %>%
  group_by(AGE) %>%
  summarise(MEAN_BY_AGE = mean(AVERAGE_RATING),
            MEDIAN_BY_AGE = median(AVERAGE_RATING),
            STDEV_BY_AGE = sd(AVERAGE_RATING))

boxplot(religionData$AVERAGE_RATING~religionData$AGE)

religionData %>%
  group_by(HOUSEHOLD_SALARY) %>%
  summarise(MEAN_BY_SALARY = mean(AVERAGE_RATING),
            MEDIAN_BY_SALARY = median(AVERAGE_RATING),
            STDEV_BY_SALARY = sd(AVERAGE_RATING))


boxplot(religionData$AVERAGE_RATING~religionData$HOUSEHOLD_SALARY)

religionData %>%
  group_by(US_REGION) %>%
  summarise(MEAN_BY_REGION = mean(AVERAGE_RATING),
            MEDIAN_BY_REGION = median(AVERAGE_RATING),
            STDEV_BY_REGION = sd(AVERAGE_RATING))


boxplot(religionData$AVERAGE_RATING~religionData$US_REGION)

religionData %>%
  group_by(GENDER) %>%
  summarise(MEAN_BY_GENDER = mean(AVERAGE_RATING),
            MEDIAN_BY_GENDER = median(AVERAGE_RATING),
            STDEV_BY_GENDER = sd(AVERAGE_RATING))

boxplot(religionData$AVERAGE_RATING~religionData$GENDER)

religionData %>%
  group_by(AGE,GENDER) %>%
  summarise(MEAN_BY_AGE_GENDER = mean(AVERAGE_RATING),
            MEDIAN_BY_AGE_GENDER = median(AVERAGE_RATING),
            STDEV_BY_AGE_GENDER = sd(AVERAGE_RATING))

ggplot(religionData, aes(x=AVERAGE_RATING))+ geom_histogram()



```

